---
title: "Writing reports in Quarto (or R markdown)"
format: revealjs
---

## Basics: Quarto and Rmarkdown

::: {.incremental}

- Quarto and R Markdown a special kind of scripts where text and computer code can be combined to generate reports. 
- Under the hood, a report generator is converting code and plain text to an output format such as *html*, *pdf* or *docx* (more formats are available).
- Quarto is a new, [well documented](https://quarto.org/) format that gives extra flexibility, but also requires installation of extra software.
- R Markdown is also, see e.g. [R Markdown](https://rmarkdown.rstudio.com/), [R Markdown: The Definitive Guide](https://bookdown.org/yihui/rmarkdown/)

:::

## Code execution and the environment

::: {.incremental}
- When a quarto or Rmarkdown file is "knitted" or compiled, the source file looks for additional files in the same directory as the source file is saved.
- Working in a RStudio projects makes it easy to work with the report interactively as you can use relative paths.
- [RStudio has an excellent guide to its project feature](https://support.rstudio.com/hc/en-us/articles/200526207-Using-RStudio-Projects)
:::

## RStudio projects

- **A projects is a collection of settings together with a root directory.**
- This means that you will be able to work with relative paths. 
- If reading a `csv` file using relative paths, your code will look like this from a project.


```{r}
#| eval: false
#| echo: true
dat <- read_csv("./data/my-data.csv")
```


## Absolute paths

- If your are using absolute paths, reaching the same operation could look like this

```{r}
#| eval: false
#| echo: true
dat <- read_csv("C:/Users/Daniel/Dropbox/Some-folder/a-project/data/my-data.csv")

```

## A basic structure for projects

- RStudio projects helps you create good habits for reproducible analysis as all analyses are conducted within a stand-alone folder structure. Your data and scripts can be shared.

- Use a basic structure for all projects:

```{}
My-project
        |
        |-.Rproj        (The project settings)
        |--/data        (Contains all data needed for your analysis)
        |--/R           (Contains all scripts/R-files)
        |--/output      (Collection of all output files)


```

## Writing in Quarto/R Markdown

- The basic syntax in quarto/R Markdown files is markdown. Markdown makes it easy to format text without point-and-click as all formatting can be added with syntax, example:

```
This text is an example of the markdown syntax which includes **bold**, *italic*,
^super^ and ~subscript~ and ~~striketrough~~
```

Resulting in:

> This text is an example of the markdown syntax which includes **bold**, *italic*,
^super^ and ~subscript~ and ~~striketrough~~

## Markdown - More on text formatting  (html)

:::: {.columns}

::: {.column width="40%"}

```


We can also use html for <sub>subscripts</sub>
and <sup>superscript</sup>


```


:::


::: {.column width="60%"}

We can also use html for <sub>subscripts</sub>
and <sup>superscript</sup>

:::

::::


## Markdown - More on text formatting  (latex)

:::: {.columns}

::: {.column width="50%"}

```

Mathematical formulas can be added inline with $y_i \sim \operatorname{Normal}(\mu_i, \sigma)$

or as a block, using

$$
y = \beta_0 + \beta_1 \times x_1 + \epsilon
$$


```


:::


::: {.column width="50%"}


Mathematical formulas can be added inline with $y_i \sim \operatorname{Normal}(\mu_i, \sigma)$

or as a block, using

$$
y = \beta_0 + \beta_1 \times x_1 + \epsilon
$$


:::

::::

## Markdown lists

:::: {.columns}

::: {.column width="40%"}

```
A list can be unordered and ordered

* Item A
    + Item x

    
1. Item 1
2. Item 2
    i) Item i


```

:::


::: {.column width="60%"}
A list can be unordered and ordered

* Item A
    + Item x

<br>

1. Item 1
2. Item 2
    i) Item i
 
:::


::::

## Markdown tables

:::: {.columns}

::: {.column width="40%"}

```

| Left |Center|Right| 
| :--- |:---: |---:|
|Text | Text| Text| 
|35| 3| 999|

: A table caption


```

:::


::: {.column width="60%"}


| Left |Center|Right| 
| :--- |:---: |---:|
|Text | Text| Text| 
|35| 3| 999|

: A table caption {.striped .hover}

:::


::::


## Adding images

:::: {.columns}

::: {.column width="40%"}

```
![](img/inn-logo.svg)

```

:::


::: {.column width="60%"}


![](img/inn-logo.svg)


:::


::::

## Cross-referencing

- With cross-referencing it is possible to reference images, tables, equations etc.
- A cross reference requires an **identifier**


:::: {.columns}

::: {.column width="40%"}

```
![](img/inn-logo.svg){#fig-inn}

See [@fig-inn] for an example ...



```

:::


::: {.column width="60%"}


![Høgskolen i Innlandet logo](img/inn-logo.svg){#fig-inn}

See [@fig-inn] for an example ...



:::


::::




## Code chunk options

- Output from code chunks can be tables, figures or text used in the report. 
- Code chunk options lets you specify, for example
    + Figure captions, figure short-captions
    + If code, messages or warnings should be shown
    + Labels used for cross-referencing
    + Caching (saving computations of a code block if unchanged)
    
    
## Code chunk options syntax

:::: {.columns}

::: {.column width="40%"}

```

#| echo: false
#| message: false
#| warning: false
#| label: fig-example
#| fig-cap: "An example figure"


library(exscidata); library(tidyverse)

cyclingstudy %>%
        ggplot(aes(weight.T1, height.T1)) + geom_point()


``` 




:::


::: {.column width="60%"}

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-example
#| fig-cap: "An example figure"


library(exscidata); library(tidyverse)

cyclingstudy %>%
        ggplot(aes(weight.T1, height.T1)) + geom_point()



```



:::


::::

## Inline code

- Code may be included inline to include code generated outputs

```{r}
#| eval: false
a_variable <- 3.14

```

- To include the variable in the text:

```{r}
#| eval: false
The variable will be displayed here `r a_variable` 
```


## Bibliographies

- Bibliographies/Citations may be added to reports using a external bibliography file. A simple format is bibtex
- Pubmed entries can be searched using [TexMed](https://www.bioinformatics.org/texmed/)
- In the visual editor, bibliographies can be easily created

```{r}
#| eval: false
#| echo: true
---
title: "My document"
bibliography: "resources/bibliography.bib"
---

```







## Where to find help?

- The  [quarto documentation](https://quarto.org/docs/authoring/markdown-basics.html) is really good.
- A lot of examples on quarto is available trough google!
    
 See the.



